---
title: "PSTAT 126 Final Project: Evan Bright"
output: pdf_document
date: "2026-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1) Select a Random Sample

```{r}

data <- read.csv("/users/evanbright/desktop/Diamonds Prices2022.csv")

set.seed(123)

diamonds <- data[sample(1:nrow(data), 1000), ]

head(diamonds)

```
2) Describe all variables:

```{r}

# Summary of the sample data
summary(diamonds)

# Structure of the sample data
str(diamonds)

```

Interpretation:

Carat Variable (Quantitative):

    - Description: The carat variable describes the weight of a diamond with 
    1 carat equalling 0.2 grams.
    - Range: 0.23 - 2.32
    - Median: 0.71
    - Mean: 0.7962
    
Depth Variable (Quantitative):

    - Description: The depth variable describes the total depth percentage 
    of the diamond, explaining the proportion between length and width.
    - Range: 57.5 - 69.6
    - Median: 61.9
    - Mean: 61.79
    
Table Variable (Quantitative):

    - Description: The table variable describes the total table percentage
    of the diamond, explaining the proportion between the width of the table
    and the width of the entire diamond.
    - Range: 52.0 - 66.0
    - Median: 57.0
    - Mean: 57.41

Price Variable (Quantitative):

    - Description: The price variable describes the price in dollars of the 
    diamond.
    - Range: 368.0 - 18706.0
    - Median: 2522.0
    - Mean: 3935.9
    
x, y, and z Variables (Quantitative):

  x:
    - Range: 3.92 - 8.57
    - Median: 5.72
    - Mean: 5.73
  
  y:
    - Range: 3.96 - 8.52
    - Median: 5.72
    - Mean: 5.74
    
  z:
    - Range: 0 - 5.280 (Seems to be a outlier of 0 in the data)
    - Median: 3.54
    - Mean: 3.54
    
  - Description: These variables represent the dimensions of the diamond
      with length, width, and depth in millimeters.

    
Cut Variable (Categorical):

    - Description: The cut variable describes the quality of the diamonds
    proportions and ability to reflect light
    - Levels: Fair, Good, Very Good, Premium, Ideal

Clarity Variable (Categorical):

    - Description: The clarity variable describes how flawless the diamond is
    both externally and internally with I1 having the most flaws and IF having
    the least flaws
    - Levels: I1, SI2, SI1, VS2, VS1, VVS2, VVS1, IF
  

Color Variable (Categorical):

    - Description: The color variable describes how much color the diamond has
    with colorless diamonds (D) being the best.
    - Levels: J, I, H, G, F, E, D

Histograms for the Continuous Variables:

```{r, include = FALSE}
library(tidyverse)
```

```{r}

# Carat Plot

carat_plot <- ggplot(diamonds, aes(x = carat)) + 
              geom_histogram(fill = "skyblue", color = "black", bins = 25) + 
              labs(title = "Distribution of Diamond Carats",
              x = "Carats",
              y = "Count") + theme_minimal()
carat_plot
```

Based off the histogram below we can see that it has a strong
right skew with the majority of the diamonds being under 1 carat. We can see
big spikes around 0.25, 0.5, 0.75, 1, and 1.5. This makes sense because in 
the diamond industry its more desirable to have a solid carat number. A 0.83
or 0.97 carat diamond would be less desirable than a solid 1.0 carat diamond.

```{r}

# Depth Plot:

depth_plot <- ggplot(diamonds, aes(x = depth)) + 
              geom_histogram(fill = "lightgreen", color = "black", bins = 25) + 
              labs(title = "Distribution of Diamond Total Depth Percentage",
              x = "Total Depth Percentage",
              y = "Count") + theme_minimal()
depth_plot

```
Based off the depth histogram we can see a relatively normal distribution with
most of the values being between 60 and 64 total depth percentage. This makes
sense because the most desirable depth percentages is around 60 to 63% which 
follows our plot.

```{r}

# Table Plot:

table_plot <- ggplot(diamonds, aes(x = table)) + 
              geom_histogram(fill = "darkorchid3", color = "black", bins = 15) + 
              labs(title = "Distribution of Table Percentage",
              x = "Table Percentage",
              y = "Count") + theme_minimal()
table_plot

```
Based off the table percentage plot we see a slight right-skew in the 
distribution most values being between 54 and 60 %. This makes sense because
the more desirable table percentages are between 53 - 58% which follows our
plot.

```{r}

# Price Plot:

price_plot <- ggplot(diamonds, aes(x = price)) + 
              geom_histogram(fill = "coral", color = "black", bins = 25) + 
              labs(title = "Distribution of Diamond Prices",
              x = "Price ($)",
              y = "Count") + theme_minimal()
price_plot

```
Based off the price plot we can see a strong right-skew with the majority of
the diamonds prices being under 5000$. The extremely high priced diamonds are
much rarer than the less priced ones.


```{r}

# X Plot (Length):

x_plot <- ggplot(diamonds, aes(x = x)) + 
              geom_histogram(fill = "slateblue", color = "black", bins = 20) + 
              labs(title = "Distribution of Diamond Length",
              x = "Length (mm)",
              y = "Count") + theme_minimal()
x_plot

```

The X plot shows us a pretty spread out distribution with a very slight bell
curve and a spike of diamonds around 4.5 mm in length.

```{r}

# Y Plot (Width):

y_plot <- ggplot(diamonds, aes(x = y)) + 
              geom_histogram(fill = "palegreen4", color = "black", bins = 20) + 
              labs(title = "Distribution of Diamond Widths",
              x = "Width (mm)",
              y = "Count") + theme_minimal()
y_plot

```

The Y plot shows us another slight bell curve with a large number of diamonds
around 4.5 mm width similar to the X plot.

```{r}

# Z Plot (Depth):

z_plot <- diamonds %>% filter(z > 0) %>%
              ggplot(aes(x = z)) + 
              geom_histogram(fill = "lightpink", color = "black", bins = 15) + 
              labs(title = "Distribution of Diamond Depths",
              x = "Depth (mm)",
              y = "Count") + theme_minimal()
z_plot

```

The Z plot shows us a somewhat right-skewed distribution with a spike of values
at around 2.5 mm and 4 mm.


Categorical Variable Plots:

```{r}

# Cut Plot:

diamonds$cut <- factor(diamonds$cut, 
                       levels = c("Fair", "Good", "Very Good", 
                                  "Premium", "Ideal"))

cut_plot <- ggplot(diamonds, aes(x = cut, fill = cut)) +
            geom_bar() + 
            labs(title = "Distribution of Diamond Cuts", 
            x = "Cut", 
            y = "Count") + theme_minimal() +
            scale_fill_brewer(palette = "Greens")
cut_plot

```
The cut bar plot shows us a strong left-skewed distribution with the
majority of diamonds having the ideal cut and the least having the fair cut.
This tells us that the better cut diamonds are more prevalent in our random
sample of the dataset.

```{r}

# Clarity Plot:

diamonds$clarity <- factor(diamonds$clarity, 
                           levels = c("I1", "SI2", "SI1", "VS2", "VS1",
                                      "VVS2", "VVS1", "IF"))

clarity_plot <- ggplot(diamonds, aes(x = clarity, fill = clarity)) + 
              geom_bar() + 
              labs(title = "Distribution of Diamond Clarities",
              x = "Clarity",
              y = "Count") + theme_minimal() + 
              scale_fill_brewer(palette = "Purples")
clarity_plot

```
The clarity bar plot has a slight bell-curve shape along with a strong 
right-skew. It has a majority of diamonds with the middle range clarities, from
SI2 to VS1 and less diamonds with the lowest and highest clarity grades.

```{r}

# Color Plot:

color_plot <- ggplot(diamonds, aes(x = color, fill = color)) +
              geom_bar() + 
              labs(title = "Distribution of Diamond Colors",
              x = "Color",
              y = "Count") + theme_minimal() +
              scale_fill_brewer(palette = "Blues", direction = -1)
color_plot

```
The color bar plot shows us a slight bell-curve with the majority of diamonds
in the middle range of the color grade. The highest and lowest color grades
are the rarest in our random sample.

3) Determine if there is any correlation between variables:

Quantitative Variables:

```{r}

quant_correlation <- cor(select(diamonds, "carat", "depth", "table", "price", 
                                "x", "y", "z"))

quant_correlation

```

Quantitative Variable Correlation Matrix Interpretation: 

- Based of the correlation matrix between quantitative variables we can see that 
many of the variables have a high correlation to eachother.

- For example, carat has strong correlation with price, x, y and z. Which
would make sense because the more a diamond weighs the bigger it must be
and the more expensive it must be as well.

- x, y, and z themselves also have a very high correlation which makes sense since
they are the diamonds dimension variables and they would go hand in hand.

- Depth and table have very weak negative correlation to eachother 
and weak positive correlation with all other quantitative variables.

Categorical Variable Correlation Interpretation:

We can't use a simple correlation matrix to describe the relationship 
between quantitative and categorical variables so we must use boxplots
to explain the correlation between the two.

I will be comparing the three categorical values to four of the quantitative
variables (leaving out x, y, and z).

Carat vs Color:

```{r}
ggplot(diamonds, aes(x = color, y = carat)) + geom_boxplot()
```
From the boxplot there seems to be a slight positive correlation between carat
and color as when the color grade goes up then the carats do as well.
This means that the majority of bigger diamonds have less color to them.

Carat vs Clarity:
```{r}
ggplot(diamonds, aes(x = clarity, y = carat)) + geom_boxplot()
```
From the boxplot we can see that there seems to be a negative correlation
between carat and clarity as when clarity goes up carats go down. This means
that higher clarity diamonds seem to be more common in smaller diamonds.

Carat vs Cut:
```{r}
ggplot(diamonds, aes(x = cut, y = carat)) + geom_boxplot()
```
From the boxplot there doesn't seem to be a strong correlation between carats
and the cut of the diamond with possibly a very weak negative correlation
as the highest grade cut seems to have the smallest carats on average.

Depth vs Color:
```{r}
ggplot(diamonds, aes(x = color, y = depth)) + geom_boxplot()
```
From the boxplot there seems to be no correlation between total depth % and 
color of the diamonds.

Depth vs Clarity:
```{r}
ggplot(diamonds, aes(x = clarity, y = depth)) + geom_boxplot()
```
From the boxplot there doesn't seem to be any correlation between total depth %
and the clarity of the diamonds. 

Depth vs Cut:
```{r}
ggplot(diamonds, aes(x = cut, y = depth)) + geom_boxplot()
```
From the boxplot there seems to be a possible weak negative correlation
between depth and cut. Although it might just be the fair cut acting as an
outlier in the dataset since we know there wasn't a high number of them.

Table vs Color:
```{r}
ggplot(diamonds, aes(x = color, y = table)) + geom_boxplot()
```
From the boxplot there doesn't seem to be any correlation between total table
percentage and the color of the diamonds. 

Table vs Clarity:
```{r}
ggplot(diamonds, aes(x = clarity, y = table)) + geom_boxplot()
```
From the boxplot there doesn't seem to be any correlation between total table
% and clarity of the diamonds.

Table vs Cut:
```{r}
ggplot(diamonds, aes(x = cut, y = table)) + geom_boxplot()
```
From the boxplot there doesn't seem to be a correlation between total table %
and cut of the diamonds. Although, the ideal diamonds seem to have a smaller
average total table % compared to the other cuts.

Price vs Color:
```{r}
ggplot(diamonds, aes(x = color, y = price)) + geom_boxplot()
```
From the boxplot there seems to be a slight negative correlation between color
and price (Since D is the highest value of color). The reason this is happening
is because although D is the best color value it is being influenced
by other variables which causes it to look like J has the highest price.

Price vs Clarity:
```{r}
ggplot(diamonds, aes(x = clarity, y = price)) + geom_boxplot()
```
From this boxplot we again see something similar to the last one. Although
IF is the best clarity it seems to be on the lower end of price. This occurs
because of other variables such as carat. The medium clarity grade diamonds
are much larger than the VVS2, VVS1, and IF causing the prices to reflect that.

Price vs Cut:
```{r}
ggplot(diamonds, aes(x = cut, y = price)) + geom_boxplot()
```
From the boxplot we don't see much correlation although we can again see that
the highest grade of cut has the lowest median price. This is again because
the size of the diamonds and other factors are influencing the price.

Correlation Interpretation:

- Based on the evidence from our correlation matrix and box-plots between variables
we can see that there isn't a ton of correlation. 
- Carat, price, x, y, and z have
some clear correlation between each other which makes sense because the dimension
variables would correlate with each other along with the weight and thus the
price of the diamonds. 
- We also see a slight correlation between clarity
and carat with smaller diamonds seeming to have the highest clarity in our
sample.

4) Summary Statistics of the Multiple Linear Regression Model:

```{r}

full_model <- lm(price ~ carat + clarity + color + cut + depth +
                   table + x + y + z, data = diamonds)

summary(full_model)

```

Summary Statistics Interpretation:

- From our summary of the multiple linear regression we can see that our 
R^2 and adjusted R^2 are both very high which means that the predictors explain
almost all (~93%) of the variability in price.
- We can see that almost all of the predictors have a very small p-value meaning
that they are statistically significant in predicting the price of the diamonds.
The only predictors that aren't stastically significant on the 0.05 level are
colorF, cutGood, cutVeryGood, cutPremium, table, y, and z.
- We can see that as clarity grades go up they seem to have higher positive
value intercepts meaning that the higher clarity the bigger effect on price.
- When it comes to color we see that the baseline (color D) has the best
positive effect on price with it getting more and more negative as it goes down
the color grades.
- Cut doesn't seem to have a specific direction that increases the price more
compared to the baseline but the ideal cut does have the largest positive
effect on price.
- Overall, from the summary statistics we can see that the effect of carat
dominates the price of the diamond with a positive value of 14152.49 which
is massive compared to the other coefficients. We also see that the length (x) 
variable has the largest negative effect on the price with a coefficient of 
-2845.46.

5) Comments on the first part:

The data within our random sample of the diamond dataset was about what I was
expecting. We saw that the weight of the diamond has the largest effect on the
price of the diamonds. We also saw that the better the cut, clarity, and color
of the diamond the seemingly higher price it would be, for the most part. There
didn't seem to be any major surprises when it came to looking at the plots of 
the variables along with their correlation.


Part 2: Simple Linear Regression

1) Create a Simple Linear Regression Model with Carat Predictor and Price
Response:

```{r}
slr_model <- lm(price ~ carat, data = diamonds)
slr_model
```

2) Examine Summary Statistics and Interpret the Simple Linear Regression Model:

```{r}
summary(slr_model)
```

- Model: $\widehat{\text{price}} = -2456.41 + 8028.56(\text{Carat})$

- Hypothesis Testing: 
    - $H_0: \beta_1 = 0$
    - $H_1: \beta_1 \neq 0$
    - t-value: 75.97 (large)
    - Since p-value < 0.05 we would reject the null hypothesis ($H_0$)
    the carat predictor has a statistically significant effect on the 
    price of the diamonds
    
Carat Predictor Interpretation:
- When the carat predictor goes up by 1 unit it causes the price of the
diamonds to go up by 8028.56$ on average which is very large

Residuals Interpretation:
- Min: -5942.7; Max: 8572.6; Median: 0.8.
- The median of the residuals is near 0 while the endpoints are very small
and very large which may show us evidence of heteroskedasticity or 
non-constant variance which we will have to check in a residuals plot

Residual Standard Error Interpretation:
- The residual standard error is 1529 which means that on average
the price prediction of the diamonds is off by 1529$.

$R^2$ and $R_{adj}^2$ Interpretation:
- We can also see that it still has a very high $R^2$ (0.8526) and 
$R_{adj}^2$ (0.8524) which tells us that just the carat predictor explains about 
85% of the variability in price just by itself.

F-statistic Interpretation:
- The F-statistic is very large (5771 on 1 and 98 DF) and has a very small 
p-value < 0.05 which means that the entire model is very statistically 
significant (in this case just with the carat predictor).

Confidence Interval:

```{r}
interval_data <- data.frame(carat = 1)
predict(slr_model, newdata = interval_data, interval = "confidence")
```

The model estimates that at a 95% confidence level the price of a 1 carat 
diamond would be between 5468.29-5676.02$, with a mean price of
5572.15$. 
We have 95% confidence that the our sample of 1 carat diamonds would be within
this confidence interval.

Prediction Interval:

```{r}
predict(slr_model, newdata = interval_data, interval = "prediction")
```

The prediction interval is much wider with a range of 2570.08 to 8574.23$ at a
95% prediction level, but has the predicted mean diamond price of 5572.15$ which
is the same as the previous.
The reason that the interval is much wider is because the prediction interval
must take into account uncertainty in estimation and variability among the
individual diamonds.

Plots of Simple Linear Regression Model:

```{r}
plot(slr_model, which = 1)
plot(slr_model, which = 2)
```

Based on the Residuals vs Fitted plot we can see that the best fit line has a
bit of a curved U-shape which suggests that carat and price have a non-linear
relationship. 
We can also see clear evidence of heteroskedasticity or non-constant variance
as the model seems to be less accurate with larger and more expensive diamonds 
compared to the smaller and cheaper ones. Although, there are a lot more points
close to the line before 5000 compared to after so most of the diamond 
population seems to fit closer to the best fit line.
Similarly, when we look at the Q-Q plot we can see that the tails fade off
significantly from the line especially at the more expensive values. This
suggests that the model under-predicts the price of higher diamonds which is
likely because we saw with the Residuals vs Fitted plot that non-linearity
is prevalent.

3) Test Assumptions and Transform the Response Variable and Predictors:

- Linearity and Constant Variance Assumption:

Based on our residuals vs fitted plot in the previous exercise we could see
evidence of non-linearity and non-constant variance within the model so we 
transform the data with the log function to try and counteract this:

```{r}
# before log transformation
plot(slr_model, which = 1)

# after log transformation
transformed_slr <- lm(log(price) ~ log(carat), data = diamonds)
plot(transformed_slr, which = 1)
``` 
Normality Assumption:

Based on the previous plots we could also see evidence of non-normality
in the Q-Q plot with values towards the end beginning to fan off. We will see
if the log transformation fixed the non-normality as well

```{r}
# before log transformation
plot(slr_model, which = 2)

# after log transformation
plot(transformed_slr, which = 2)
```

Linearity: 
- After using a log function on the model we can clearly see that the
non-linearity is gone as the best fit line is horizontal and we see a random
assortment of values around the line.
- The assumption of linearity is now valid.

Constant Variance:
- After using the log function on the model we could see that the fanning of 
values on the residuals vs fitted plot were reduced and values got closer
to the best fit line with no evidence of heteroskedasticity.
- The assumption of constant variance is now valid.

Normality:
- After using the log function we can also see that the Q-Q plot now has no
points fanning off at the end other than a small few at the very beginning of 
the plot. This means that not only did the log transformation fix the 
non-linearity it also fixed the normality as well.
- Thus, the normality assumption is now valid.

This means that after the log transformation the model now follows all the
assumptions.

4) Summary of Transformed Variables:

```{r}
summary(transformed_slr)
```

Predictor Interpretation:
- After log transformation the carat predictor is still statistically 
significant to price although it's value has fallen drastically to just ~1.69, 
while the intercept value is ~8.45 which is also statistically significant.
- With the transformation it now reads that for a 1 unit increase in carat
it increases the log of price by ~1.69$.

Residuals:
- Min: -1.02; Max: 0.81; Median: -0.002
- We can see that the residual range has gotten much smaller and is much
closer to the median value which is located at practically zero.
- This means that the transformation has reduced the variability in the data 
with values much closer to the median value at ~0.

Residual Standard Error:
- Was 1529 on the original model and is now just 0.25 on the transformed model.
- This explains that the error of the residuals is now much smaller and the 
model fits the data much more accurately.

R-squared and Adjusted R-squared:
- Original Model: $R^2$ = 0.8526 and $R_{adj}^2$ = 0.8524
- Transformed Model: $R^2$ = 0.938 and $R_{adj}^2$ = 0.9379
- This tells us that the transformed model explains more variability (about 
~94%) of the log of the price response variable.

F-statistic:
- The F-statistic is extremely large, larger than the original model, and the
p value is still very small (< 0.05)
- This means that the overall model is statistically significant, and that
the log of carat predictor explains a large majority of the variability
in the data.

5) Add Predictors To The Model And Observe If There's Improvement:

```{r, include = FALSE}
# transformed baseline slr model (adjusted R-squared: 0.9379)
summary(transformed_slr)

# multiple linear regression model (adjusted R-squared: 0.9845)
mlr_model <- lm(log(price) ~ log(carat) + cut + clarity + color + x
                , data = diamonds)
summary(mlr_model)

```

Starting with our simple linear regression model with a log transformation
on both our response variable (price) and predictor (carat) we already had
an adjusted R-squared of 0.9379. This is already a very good model explaining
almost all of the variability in the data just with one predictor.

The first variable that I added to the model was cut. With the additional
predictor in the model the adjusted R-squared improved to 0.9424. Each cut
level was statistically significant with p-values less than 0.05, so
it improves upon the model. Since it improved the adjusted R-squared and is 
statistically significant we kept the cut predictor in the model.

The next variable that I added to the model was clarity. With this predictor
added to the previous model we got an increase in the adjusted R-squared up
to 0.9682. Additionally, each level of clarity was statistically significant
with p-values less than 0.05, thus it improves the model. Since it improved
adjusted R-squared and all levels were statistically significant we kept the
clarity predictor in the model.

We then added the final categorical predictor to the model, color. After adding,
it improved the adjusted R-squared to 0.9838. Each level of color is also 
statistically significant with p-values less than 0.05, improving the model. 
With the improvement of the adjusted R-squared and significance of all levels,
we kept the clarity predictor in the model.

The next variable I added was depth. The adjusted R-squared saw no increase
staying at 0.9838 and the predictor was not statistically significant with a 
large p-value of 0.685. With no effect on adjusted R-squared and a lack of
significance we did not keep the depth predictor in the model.

Then we added the table variable to the model. Similarly to the previous,
we saw no change in the adjusted R-squared staying at 0.9838 and the table
predictor also had a large p-value of 0.966 meaning that it isn't statistically
significant and doesn't leave an effect on the model. Since there was no effect
on the adjusted R-squared and no statistical significance we did not keep the
table predictor in the model.

Moving on to the dimensional variables we started with adding the x predictor.
After adding it ended up improving the adjusted R-squared to a value of 0.9843
and was statistically significant with a p-value less than 0.05. With the
improvement of the adjusted R-squared along with the statistical significance
we kept the x predictor in the model.

We then moved on to the y predictor and added it to the model. After doing so
we saw no change in the adjusted R-squared and the predictor was not
statistically significant at the 0.05 level with a p-value just under 0.1. With
the lack of improvement on the adjusted R-squared and the insignificance of 
the predictor we did not keep it in the model.

Finally we added the last predictor, z, to the model. With the addition of
the z predictor we saw no change in the adjusted R-squared with the model 
ending at a adjusted R-squared value of 0.9843. The z predictor was also seen 
to not be statistically significant with a p-value of 0.31. With no 
improvement in the adjusted R-squared and no statistical significance we did
not keep the z predictor in the model.

The final multiple linear regression model that we ended up with involved
6 of the predictors. These were the log(carat), cut, color, clarity and x 
predictors and didn't include depth, table, and the y predictors. The final
adjusted R-squared improved from the simple linear regression model value of
0.9379 to a value of 0.9843 which means they explain about 98.4% of the
variability in the sampled data. This is a very high value and thus we have 
almost explained the entirety of the variability in our sample.

Multiple Linear Regression Model:

log(price) ~ log(carat) + cut + color + clarity + x

6) Comments:

When adding predictors everything went according to what I had assumed would
happen. I was a bit surprised that the table and depth predictors had no 
effect on the model which must be due to the other predictors explaining
the same variation in the sample as they do.

I wasn't surprised with the lack of improvement from the y and z predictors
as there was a high correlation between them with not only the x predictor
but also the carat predictor which dominates the price response variable.

I had also assumed that the categorical predictors, cut, color and clarity
would have a significant positive effect on the model which was true in this
case.

Overall, when constructing the multiple linear regression model by adding 
the predictors one at a time we were able to improve our adjusted R-squared
and add numerous predictors that helped improve the model while being
statistically significant.


Part 3: Part 2 Continuation

1) AIC and BIC Methods

```{r}
# use the aic method with a forward direction
aic_method <- step(full_model, direction = "both")

# the aic method only removes the y predictor

aic_mlr_model <- lm(log(price) ~ log(carat) + x + z + depth + table +
                   cut + clarity + color, data = diamonds)

summary(aic_mlr_model)
```

AIC Method Interpretation:

- From the AIC method we get a final model that uses the carat, clarity,
color, cut, depth, table, x, and z variables.
- It adds all variables except the y (width) variable.
- It likely removed the y variable since it is highly correlated with the other
dimension variables as well as the carat variable and in this case is redundant 
with the addition of the other dimension variables such as x and z in addition
to the carat predictor.
- The AIC of the full model with all predictors is 13973.61 and with the
removing of the y predictor the AIC value becomes 13973.18 which is a very
minuscule difference.
- The adjusted R-squared of the AIC method multiple linear regression model
is 0.9845 which is just minorly larger than the 0.9843 value of adjusted
R-squared we got when adding the predictors ourselves one by one.

Carat Interpretation:
- The log(carat) is still the most significant predictor with a coefficient
of 1.447, dominating the effect on the price of the diamonds.
- A 1 unit increase in log(carat) increases the log(price) by 1.447$.

Dimension Variable Interpretation:
- X has a positive effect on log(price) of 0.288.
- Z has a negative effect on log(price) of -0.08.

Cut Variable Interpretation:
- All cut levels have a positive effect on log(price) with cutIdeal having
the largest with 0.2 meaning that when a diamond has an ideal cut the log(price)
goes up by 0.2.
- This means the baseline (cutFair) has the smallest effect on the response.
- All cut levels are statistically significant with p-values < 0.05.

Clarity Variable Interpretation:
- All clarity levels have a positive effect on log(price) with clarityIF being
the largest with a coefficient value of 1.06.
- This ends up being the second most influential predictor on the log(price)
response variable which means that clarityIF has a major effect on the 
price of the diamonds
- This means that the baseline (I1) has the smallest effect on the response.
- All clarity levels are statistically significant with p-values < 0.05.

Color Variable Interpretation:
- All color values have a negative effect on log(price) with colorJ having the
largest negative effect on the response.
- This means that the baseline (colorD) has the largest positive effect on
the response.
- All color variables are statistically significant with p-values < 0.05.

Table Variable Interpretation:
- The table variable has a very small coefficient of 0.002 which suggests
that it doesn't have much of an effect on the response.
- It is also not statistically significant with a p-value > 0.05.

Depth Variable Interpretation:
- The depth variable has a very small coefficient of 0.016 which suggests that
it has minimal effect on the response.
- Although, it is statistically significant with a p-value < 0.05.


2) Detect Multicollinearity using VIF:

```{r}
library(car)
vif(aic_mlr_model)
```

When using the VIF function on the multiple linear regression made using
the AIC method we get 2 predictors with VIF values larger than 10 which are
log(carat) and x. We also see that the z predictor has a VIF value of 5 which
is concerning as well.

Since x has the largest VIF value of ~12 we will remove it from the multiple
linear regression model and take the VIF function on that model.

```{r}
new_aic_mlr_model <- lm(log(price) ~ log(carat) + cut + color + clarity + z +
                      depth + table, data = diamonds)

summary(new_aic_mlr_model)

vif(new_aic_mlr_model)
```

After using the VIF function again we see that log(carat) and the z predictor
have VIF values just slightly above 5. When checking how each predictor affects
the model we can see that log(carat) explains the majority of the variation
in the sample in addition to being statistically significant. The z predictor
on the other hand adds nothing to the adjusted R-squared while also not being
statistically significant. 

Thus, I decided to remove the z predictor from the model and ran the VIF
function one more time.

```{r}
final_mlr_model <- lm(log(price) ~ log(carat) + cut + clarity + color +
                        table + depth, data = diamonds)
summary(final_mlr_model)

vif(final_mlr_model)
```

We can now see that none of the predictors have VIF values above or even near 5.
This means we have detected and successfully removed all signs of
multicollinearity in the sample data, while also keeping our adjusted R-squared
around the same level as before. 

Thus, for my final multiple linear regression model I decided to take the model
that we used the AIC method to remove overfitting on as well as the VIF function
to remove all signs of multicollinearity. Finally, I ended up removing the table
and depth predictors myself as they both had no effect on the adjusted R-squared
and neither were statistically significant. I felt that there was no need to 
include them in the final model.

This gives us an ultimate final multiple linear regression model of:

$$\log\text{(price)} = \beta_0 + \beta_1 \log\text{(carat)} + \beta_2 \text{(cut)} 
+ \beta_3 \text{(color)} + \beta_4 \text{(clarity)} + \epsilon$$

where in our sample data: 

$\beta_0$ = 7.83 and $\beta_1$ = 1.89 with the other predictors varying with
their $\hat\beta$ as they depend on which level is present.

```{r}
final_model <- lm(log(price) ~ log(carat) + cut + color + clarity, 
                        data = diamonds)
summary(final_model)
```


3) Confidence and Prediction Intervals:

```{r}
# create a point from the model

prediction_point <- data.frame(
  carat = 1,
  cut = "Ideal",
  clarity = "IF",
  color = "D"
)

# create a 95% confidence interval for the mean log(price)
predict(final_model, newdata = prediction_point, interval = "confidence")

# create a 95% prediction interval for a future log(price)
predict(final_model, newdata = prediction_point, interval = "prediction")
```

For the confidence and prediction intervals I decided to make a point with
a carat value of 1 with each categorical predictors best value.

In this case those values are cutIdeal, clarityIF, and colorD which would
make the most "perfect" diamond if combined.

The intervals give us a range to quantify the uncertainty we have in our 
predictions of expected average price the realistic range of the prices
for a certain diamond.

Confidence Interval Interpretation:

- The confidence interval that we've created gives us a basis of our
uncertainty in the estimated mean log(price) for all of the diamonds in the 
sampled data set that align with these point values. 
- The confidence interval has a narrower range than the prediction interval 
because it is only telling us the uncertainty in estimating the average of
the response
- The estimate of log(price) in the interval is 9.065 for diamonds that are
1 carat, with ideal cut, clarity IF, and color D.
- The bounds for the interval are 9.016 - 9.113
- Thus, we are 95% confident that the true mean log(price) of diamonds
that have these point values reside between the 9.016 - 9.113 range.
- When we convert the log(price) back to an untransformed price value we
get that the confidence interval is 8233.78 - 9072.47$, with an estimated
mean value of 8647.28$.

Prediction Interval Interpretation: 

- The prediction interval we've created gives us the range for the true price
of a future diamond with these same characteristics. 
- The model predicts that a future diamond that follows these previous point 
values, will have a log(price) value of about 9.0645.
- The range of the prediction interval is 8.807 - 9.3218
- This interval is quite a bit larger than the confidence interval because
it reflects the natural variability between diamonds, the uncertainty of the
model, as well as random noise in the price.
- Thus, we are 95% confident that a future diamond with these given traits
will have a log(price) value between 8.807 and 9.3218.
- When converting from log(price) to price we get interval values 
equivalent to 6680.85 - 11179.09$ and an estimated mean price value of 
8647.28$, the same as the confidence interval.

4) Summary of the Report:

Part 1 Summary:

- In part 1, we created a random sample of 1000 diamonds from the diamonds data 
set. We gave descriptions of each variable in the data set including price, 
carat, table, depth, x, y, z, cut, color and clarity. We further explained the
variables by giving a visual representation of each and explaining their
distribution based off of those plots. This gave us a lot of information
about each variable and helped us get to a final multiple linear regression
model by the end of the project. We were also asked to check the correlation
between these variables using a correlation matrix for the quantitative 
variables and using box-plots for the categorical variables. We found that
many of the variables had strong correlation to each other, mainly carat, price,
x, y, and z. Finally, we put all of our observed variables into a multiple
linear regression model and got the summary statistics for that given model. 
When looking at the summary statistics of the model we noticed that the carat
predictor dominated with it's massive effect on the response variable (price).
After observing and describing the variables of the dataset we
decided to use price as our response variable and carat as our strongest
predictor that we would use in our simple linear regression model that we 
ended up making in part 2.

Part 2 Summary:

- Moving on to part 2 of the project we focused a majority of our time
on the simple linear regression model with price as our response variable
and carat as our $\beta_1$ predictor. We observed the summary statistics on
this simple linear regression model and saw that just the carat predictor
by itself explained almost all of the variation in the response variable which
is what we expected based on our observations of the predictors. When checking
the linear regression model assumptions we noticed a couple of problems that
would need to be fixed before claiming our final model. We observed both 
non-linearity and heteroskedasticity (non-constant variance) as well as slight
effects of non-normality in our simple linear regression model. To find 
linearity and constant variance in our model we decided to take a log 
transformation of both the price response variable as well as the carat 
predictor. This transformation ended up fixing our issues and thus we had a 
simple linear regression model that followed all the assumptions as it's now 
linear, has constant variance, and is normally distributed. Once we had this
complete simple linear regression model we decided to start adding additional
predictors to improve the model. We ended up adding all the categorical
values (cut, color, and clarity) as well as the x (length) predictor.
This caused our adjusted R-squared to increase meaning that we improved upon
our simple linear regression model.

Part 3 Summary:

- In part 3 we further tried to improve our multiple linear regression model by
using the aic method. Specifically, we used the full model with all predictors
and went with the "both" direction. Using this method, we ended up getting
a multiple linear regression model with the smallest AIC value which had all
of the predictors excluding just the y (width) predictor. This solved our
problem of over-fitting so we moved on to our detection of multicollinearity
within the model. By using the VIF function we were able to detect strong
multicollinearity between the x, z, and carat predictors. Using these values
we ended up removing both the x and z predictors from the model and thus had
a multiple linear regression with no over-fitting and no multicollinearity. 
Finally, I also decided to remove the table and depth predictors from the model
as even though they remained during the aic method I saw no reason to keep them
as they didn't improve our adjusted R-squared and were not statistically
significant. Now that we had our final multiple linear regression model
that best explained the random sample we made of the diamonds dataset we
created both a confidence and prediction interval for diamonds with a specific
set of values for each predictor. Overall, after all of these processes
we ended up with our "perfect" multiple linear regression model that captured
the data the best while explaining the majority of variation within the 
price of diamonds while following our assumptions and containing no over-fitting
or multicollinearity. 





